From physical principles to ???

### 2015-12-14 2015CompleteDerivation/completeDerivation.tex
Refined composite/classical state definitions

Given that scientific reductionism (i.e. the idea of reducing physical systems and interactions to the sum of their constituent parts in order to make them easier to study) is at the heart of fundamental physics, we should explore how to characterize a system in terms of its components. In general, this is quite a complicated thing to do, that requires intimate knowledge of the system at hand. We will study an idealized limit case: one where the components are homogeneous and infinitesimal. What we'll find is that under the assumption that the system is reducible (i.e. its state is equivalent to the state of the parts) and that each part undergoes deterministic and reversible evolution, the motion is suitably described by the standard framework of classical Hamiltonian particle mechanics.

\subsection{Composite systems}
As mentioned before, we will limit ourselves to the case where our system is homogeneous: all parts can be characterized in the same way, they are under the same external and internal forces, the set of states they can occupy is the same. For a non-homogeneous system, we would divide it in homogeneous components and consider the state of all of them. Ultimately, what homogenous means depends on context (e.g. air can be thought as homogeneous if the mixture of gasses does not change in space or in time due to phase transitions or chemical processes) and one must check that such property is conserved by time evolution.

\begin{defn}\label{reducible_state_space}
The state space $\mathfrak{C}$ for a homogeneous decomposable system is an additive abelian (i.e. commutative) group.
\end{defn}

\begin{justification}
We first claim that the state space $\mathfrak{C}$ is an additive monoid: there exist a law of composition $+ : \mathfrak{C} \times \mathfrak{C} \rightarrow \mathfrak{C}$ that takes two states and returns one that is the physical composition of the two. The domain and codomain match because the system is homogeneous. The law is commutative $\mathfrak{c}_1 +\mathfrak{c}_2 = \mathfrak{c}_2+\mathfrak{c}_1$ and associative $(\mathfrak{c}_1 + \mathfrak{c}_2) + \mathfrak{c}_3 = \mathfrak{c}_1 + (\mathfrak{c}_2 + \mathfrak{c}_3)$, as it does not matter in what order we compose the parts. There exist a zero element $\mathfrak{c} + 0 = \mathfrak{c}$ and it represent the state with no system.

We then claim it is an additive group. As we want to describe changes during the evolution, i.e. $\Delta \mathfrak{c}$, we introduce an inverse $\mathfrak{c}_2$ for each element $\mathfrak{c}_1$ such that $\mathfrak{c}_1 + \mathfrak{c}_2 = 0$. Such inverse may not represent a true physical state, yet we'll still call $\mathfrak{C}$ the state space committing an abuse of terminology. Having an inverse, $\mathfrak{C}$ is an additive (i.e. abelian) group.
\end{justification}

\subsection{Classical composite systems}
We also want to be able to express the state of the composite system in terms of the state of the parts. For example, we'll want to relate the state variables of the composite system to the state variables of the parts. This in general will depend on how much the state of the composite system "knows" about its part. For example, the position and orientation of an ideal rigid body is enough to define where all its constituents are, while the volume/pressure/temperature of an ideal gas is not enough to determine the position and momentum of all its molecules. We need to characterize the system further.

We will call a classical system one that is infinitesimally reducible. That is: it is made of arbitrarily small parts, which we call particles, and the state of the whole system is equivalent to the

We also need to be able to characterize the composite state in terms of the

% Infinitesimally decomposable and reducible
% We have structure for pieces (their state space). We can use labels to identify pieces. We just need to consider state transformations that do not change the state variable of the pieces and are independent of state variable of the pieces. Increase number of pieces is the only thing.
% State space is vector space over R

@@

We now introduce the idea that a system is decomposable, that is it is made of smaller parts. To simplify our discussion, we will always assume that our system is homogeneous: all parts can be characterized in the same way, that they are under the same external and internal forces, that the set of states they can occupy is the same. For a non-homogeneous system, we would divide it in homogeneous components and consider the state of all of them. Ultimately, what homogenous means depends on context (e.g. air can be thought as homogeneous if the mixture of gasses does not change in space or in time due to phase transitions or chemical processes) and one must check that such property is conserved by time evolution.

@@

\begin{defn}\label{vector space}
The state space $\mathfrak{C}$ for a infinitesimally reducible system is a vector space. It admits a basis $\mathfrak{S}$ that represents the state space for the infinitesimal parts.
\end{defn}

\begin{justification}
Since the system is decomposable, the state space is an abelian group.

We now consider the set of transformations $A$ that increase or decrease the size of the system by a constant. This forms a field\footnote{Here field is intended in the abstract algebraic sense (a nonzero commutative division ring) which has no relationship to the field in the physics sense (a physical quantity with a value for each point in space).} that is isomorphic to $\mathbbm{R}$. In fact, let $a: \mathbbm{R} \rightarrow A$ be a mapping between a number and the transformation that increases or decreases the size of the system by that number; define on $A$ an addition $+: A \times A \rightarrow A$ and a multiplication $*: A \times A \rightarrow A$ such that $a(x) + a(y) = a(x+y)$ and $a(x) * a(y) = a(x*y)$, so that the sum and product of the transformation is equal to the sum and product of their respective factors. $A$ is a field, as its sum and multiplication have all properties of $\mathbbm{R}$ which is a field. $a$ is an homomorphism by definition and is invertible by the very definition of $A$, each element fully identified by its constant. Therefore $A$ and $\mathbbm{R}$ are isomorphic.

We now claim that the state space $\mathfrak{C}$ is a vector space over $\mathbbm{R}$. The abelian group $\mathfrak{C}$ can be extended with the operations defined by $A$, as each element $a \in A$ is a map $a : \mathfrak{C} \rightarrow \mathfrak{C}$. The map has the following properties: $(a_1 + a_2) \mathfrak{c} = a_1 \mathfrak{c} + a_2 \mathfrak{c}$, increasing the size of the system by the sum of two constant is the same as combining the separate increases, and $a (\mathfrak{c}_1 + \mathfrak{c}_2) = a \mathfrak{c}_1 + a \mathfrak{c}_2$, increasing the size of the total system is the same as the combination of the increased parts. $\mathfrak{C}$ is a module over $A$, which is a field and isomorphic to $\mathbbm{R}$. We can then think of $\mathfrak{C}$ as a vector space over $\mathbbm{R}$.
\end{justification}

>>>>

Given that scientific reductionism (i.e. the idea of reducing physical systems and interactions to the sum of their constituent parts in order to make them easier to study) is at the heart of fundamental physics, we should explore how to characterize a system in terms of its components. In general, this is quite a complicated thing to do, that requires intimate knowledge of the system at hand. So we simplify our problem and study an idealized case: one where the components are homogeneous and infinitesimal. What we'll find is that under the additional assumption that the system is reducible (i.e. its state is equivalent to the state of the parts) and that each part undergoes deterministic and reversible evolution, the motion is suitably described by the standard framework of classical Hamiltonian particle mechanics.

\subsection{Composite systems}
As mentioned before, we will limit ourselves to the case where our system is homogeneous: the parts are indistinguishable, the set of states they can occupy is the same, they are under the same external and internal forces. For a non-homogeneous system, we would divide it in homogeneous components and consider the state of all of them. What homogenous means depends on context (e.g. air can be thought as homogeneous if the mixture of gasses does not change in space or in time due to phase transitions or chemical processes) and one must check that such property is conserved by time evolution. Nonetheless, we can claim the following:

\begin{defn}\label{reducible_state_space}
The state space $\mathbbm{C}$ for a homogeneous decomposable system is an additive abelian (i.e. commutative) group.
\end{defn}

\begin{justification}
We claim $\mathbbm{C}$ is an additive monoid. There exist a law of composition $+ : \mathbbm{C} \times \mathbbm{C} \rightarrow \mathbbm{C}$ that takes two states and returns one that is the physical composition of the two. The domain and codomain match because the system is homogeneous. The law is commutative $\mathbbm{c}_1 +\mathbbm{c}_2 = \mathbbm{c}_2+\mathbbm{c}_1$ and associative $(\mathbbm{c}_1 + \mathbbm{c}_2) + \mathbbm{c}_3 = \mathbbm{c}_1 + (\mathbbm{c}_2 + \mathbbm{c}_3)$, as it does not matter in what order we physically compose the parts. There exist a unique zero element $\mathbbm{c} + 0 = \mathbbm{c}$ and it represent the physically empty state (i.e. no system).

We claim $\mathbbm{C}$ is an additive group. As we want to describe changes during the evolution (i.e. $\hat{\mathbbm{c}} = \mathbbm{c} +\Delta \mathbbm{c}$) we introduce an inverse $\mathbbm{c}_2$ for each element $\mathbbm{c}_1$ such that $\mathbbm{c}_1 + \mathbbm{c}_2 = 0$. The inverse may not represent a proper physical state, yet we'll still call $\mathbbm{C}$ the state space committing an abuse of terminology. The inverse $- : \mathbbm{C} \rightarrow \mathbbm{C}$ is physically unique and well defined.
\end{justification}

\subsection{Classical composite systems}
We also want to be able to express the state of the composite system in terms of the state of the parts. For example, we'll want to relate the state variables of the composite system to the state variables of the parts. This in general will depend on how much the state of the composite system "knows" about its parts. For example, the position and orientation of an ideal rigid body is enough to define where all its constituents are, while the volume/pressure/temperature of an ideal gas is not enough to determine the position and momentum of all its molecules. Therefore we need to characterize the system further.

We will call a classical system one that is infinitesimally reducible. That is: it is made of arbitrarily small parts, which we call particles, and the state of the whole system is equivalent to the sum of the states of the parts. Given the state space $\mathbbm{S}$ for the parts, giving a composite state $\mathbbm{c}$ is equivalent to give the amount of particles that are present in each state. Since the system is infinitesimally reducible, such amount can be increased or decreased infinitesimally and will be a real number. In other words: the classical state is a real valued distribution over $\mathbbm{S}$. This intuitive picture is unfortunately not suitable to be generalized to different contexts, so we develop another.

Consider the following expression:
\begin{align*}
\mathbbm{c} = \sum\limits_{i \in \mathbbm{I}} a_i \mathbbm{e}(i)
\end{align*}
$\mathbbm{I}$ is best description of an infinitesimal part that a composite system can give. $\mathbbm{e}(i)$ represents a composite system made of a unit amount of particles, all prepared in the state determined by $i$. $a_i$ represents a transformation that changes the whole state, without affecting the value $i$ for any infinitesimal part. For a classical system, $i$ represents the full information about the infinitesimal part, and therefore includes the value of all state variables; the only transformation left is increasing/decreasing the amount of particles by a scalar multiple. For a different system, where $i$ does not represent the full state of the parts, $\mathbbm{e}(i)$ will be an ensemble and $a_i$ will include internal transformations. This is intuitively how we decouple the composite state into the part $i$ that describes the infinitesimal parts, and the part $a_i$ that describes the rest. Giving a full description of the state space of a decomposable system means fully characterizing these components. And we can see how this fits for the classical case as well: the state variables gives us the description for an infinitesimal part, but the amount of particles for each state is a description of the composite state.

\begin{defn}\label{classical_vector space}
The state space $\mathbbm{C}$ for a homogeneous classical (i.e. infinitesimally reducible) system is a vector space over $\mathbbm{R}$. It admits a basis isomorphic to the state space $\mathbbm{S}$ of its particles (i.e. infinitesimal parts).
\end{defn}

\begin{justification}
The state space is an abelian group as the system is homogeneous and decomposable and \ref{reducible_state_space}.

We now consider the set of transformations $A$ that increase or decrease the amount of parts of the system by a constant. We claim $A$ is a field\footnote{Here field is intended in the abstract algebraic sense (a nonzero commutative division ring) which has no relationship to the field in the physics sense (a physical quantity with a value for each point in space).} isomorphic to $\mathbbm{R}$. Consider $a: \mathbbm{R} \rightarrow A$ the mapping between a number and the transformation that increases or decreases the size of the system by that number. This transformation exists: the system is infinitesimally decomposable and the amount can be changed continuously. Define on $A$ an addition $+: A \times A \rightarrow A$ and a multiplication $*: A \times A \rightarrow A$ such that $a(x) + a(y) = a(x+y)$ and $a(x) * a(y) = a(x*y)$, so that the sum and product of the transformation is equal to the sum and product of their respective factors. $a$ is an isomorphism between $A$ and $\mathbbm{R}$. $A$ is isomorphic to $\mathbbm{R}$ as a field.

We now claim that the state space $\mathbbm{C}$ is a vector space over $\mathbbm{R}$. The abelian group $\mathbbm{C}$ can be extended with the operations defined by $A$, as each element $a \in A$ is a map $a : \mathbbm{C} \rightarrow \mathbbm{C}$. The map has the following properties: $(a_1 + a_2) \mathbbm{c} = a_1 \mathbbm{c} + a_2 \mathbbm{c}$, increasing the size of the system by the sum of two constant is the same as combining the separate increases, and $a (\mathbbm{c}_1 + \mathbbm{c}_2) = a \mathbbm{c}_1 + a \mathbbm{c}_2$, increasing the size of the total system is the same as the combination of the increased parts. $\mathbbm{C}$ is a module over $A$, which is a field and isomorphic to $\mathbbm{R}$.

Consider now the injection map $\mathbbm{e}: \mathbbm{S} \hookrightarrow \mathbbm{C}$ that for each particle state returns a composite state constituted only by a predetermined amount of particles in such state. The image of such injection is a basis for $\mathbbm{C}$. We first claim that all elements are linearly independent. Consider $\sum\limits_{\mathbbm{s}_i \in \mathbbm{S}} a_i \mathbbm{e}(\mathbbm{s}_i)$, with $a_i \in \mathbbm{R}$ and . This corresponds to a physical state with $a_i$ particles in each state $\mathbbm{s}_i$. This is equal to the empty state if and only if $a_i=0 \forall i$. This holds even if $\mathbbm{S}$ is not finite. We now claim they span. Consider $\mathbbm{c} + \sum\limits_{\mathbbm{s}_i \in \mathbbm{S}} a_i \mathbbm{e}(\mathbbm{s}_i)$, $a_i \in \mathbbm{R}$. This can always be made equal to the empty state by setting $a_i$ to be the opposite of the amount of particles of $\mathbbm{c}$ that are in state $\mathbbm{s}_i$.
\end{justification}

\subsection{Classical assumption}

@@

@@
